---
title: "TIME SERIES QUIZ 4"
author: "NEEMA NDANU"
date: "2024-07-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QUESTION 1
You have been provided with monthly sales data for a new retail store over a period of approximately 6.5 years (80 months). The data is measured in thousands of dollars.Dataset: timeseries_data.csv

### 1. Load the dataset into RStudio and plot a time series graph to visualize the monthly sales data.Provide an interpretation of any observed trends or patterns in the data.
```{r}
# Load the necessary data set
library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)
library(tseries)
library(forecast)
library(MASS)

# Read the data set
data <- read.csv("C:\\Users\\HP\\Desktop\\R- CODES\\timeseries_data.csv")

# Convert Date column to Date format
data <- data %>%
  mutate(Date = as.Date(paste0(Date, "-01")))


# Plot the time series
ggplot(data, aes(x = Date, y = Sales)) +
  geom_line(color = "blue") +
  labs(title = "Monthly Sales Over Time", x = "Date", y = "Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
The above graph shows an upward trend in sales over the entire period. This indicates that the store's monthly sales have been consistently increasing over time, which could be due to factors such as increased customer base, improved marketing strategies, or expansion of product lines.

The graph does not show any obvious seasonal patterns or fluctuations, such as periodic peaks or troughs that might align with holidays or other recurring events. This suggests that the store's sales are not heavily influenced by seasonal variations and maintain a steady increase throughout the year.

The graph does not exhibit any periods of decline or significant drops in sales. This suggests that the retail shop has a stable business environment with few disruptions indicating that the store has been resilient to external economic factors, maintaining a strong performance even in potentially challenging times.

### 2. Perform the Augmented Dickey-Fuller (ADF) test on the sales data to determine if it is stationary.Interpret the results of the test and discuss the implications for further analysis.

```{r}
# Convert the data to a time series object
sales_ts <- ts(data$Sales, start = c(2018, 1), frequency = 12)

# Perform the Augmented Dickey-Fuller test
adf_test_original <- adf.test(sales_ts)

# Print the results
cat("ADF test for original data:\n")
print(adf_test_original)
```
The Augmented Dickey-Fuller (ADF) test results indicate a test statistic of -10.674 and a p-value of 0.01. In the context of the ADF test, a more negative test statistic provides stronger evidence against the null hypothesis of non-stationarity. Therefore, with a test statistic of -10.674, the data is deemed stationary. Additionally, the p-value of 0.01 is lower than the typical significance level of 0.05, leading us to reject the null hypothesis, further supporting that the sales data exhibits stationarity.

The lag order of 4 indicates that the ADF test considered up to 4 lags in the data. However, despite the statistical indication of stationarity, a visual inspection of the time series plot reveals a clear upward trend, which is characteristic of non-stationary data. It's important to note that a p-value smaller than the chosen significance level (e.g., 0.05) indicates strong evidence against the null hypothesis.

Therefore, despite the visible upward trend in the data, statistical analysis using the ADF test suggests that the data behaves in a stationary manner. This underscores the importance of relying on statistical tests like the ADF test to make conclusions about the stationarity of time series data, rather than solely relying on visual observations. The alternative hypothesis in this context states a preference for stationarity, emphasizing the specific focus of the test in confirming stationarity.

### Apply the Box-Cox transformation to the sales data to stabilize variance.Plot the transformed data and interpret how the transformation impacts the distribution and stationarity of the data.

```{r}
# Apply the Box-Cox transformation
lambda <- BoxCox.lambda(data$Sales)
sales_transformed <- BoxCox(data$Sales, lambda)

# Add transformed data to the dataframe
data$Sales_transformed <- sales_transformed

# Print the lambda value
cat("The lambda used is:", lambda, "\n")

# Plot the transformed data
ggplot(data, aes(x = Date, y = Sales_transformed)) +
  geom_line(color = "red") +
  labs(title = "Transformed Monthly Sales Over Time", x = "Date", y = "Transformed Sales")

```
We use the BoxCox.lambda() function to determine an optimal lambda parameter value, which in this case is calculated as 0.9076083. Applying this lambda value in the BoxCox() transformation results in a time series where the variance appears more consistent.

The lambda value of 0.9076083 indicates that the Box-Cox transformation applied to the sales data is closer to a power transformation rather than a logarithmic one.

When comparing the original sales data graph to the transformed sales data graph, one notable difference is the starting points of both graphs. The original data starts at a higher value, whereas the transformed data starts closer to zero. This shift occurs because the Box-Cox transformation compresses or expands the data, altering its scale and distribution.

### Perform the ADF test on the transformed sales data to confirm stationarity.Interpret the results and compare them with those from Question 2 to assess the effectiveness of the Box-Cox transformation.

```{r}
# Perform the Augmented Dickey-Fuller test on the transformed data
adf_test_trans <- adf.test(data$Sales_transformed)

# Print the results
cat("ADF test for transformed data:\n")
print(adf_test_trans)

```
The ADF test statistic is -1.7399 and the p-value is 0.682. In the context of the ADF test, a more negative test statistic provides stronger evidence against the null hypothesis of non-stationarity. However, in this case, the test statistic of -1.7399 is not strongly negative, indicating weaker evidence against non-stationarity.

The lag order of 4 indicates that the ADF test considered up to 4 lags in the data. Furthermore, the p-value of 0.682 is much higher than the typical significance level of 0.05. This suggests that we do not have sufficient evidence to reject the null hypothesis. Therefore, based on these results, we fail to reject the null hypothesis, implying that the sales data is likely non-stationary.

Despite the statistical indications from the ADF test, a visual inspection of the time series plot reveals a clear upward trend, which is characteristic of non-stationary data. This visual observation supports the conclusion that the sales data exhibits non-stationary behavior. This contrast underscores the importance of relying on statistical tests like the ADF test to accurately assess the stationarity of time series data, as they provide quantitative measures that complement and sometimes contradict visual interpretations.

The alternative hypothesis in this context states a preference for stationarity, emphasizing the specific focus of the test in confirming stationarity.

### Compute and plot the Autocorrelation Function (ACF) for up to 20 lags of the transformed sales data.Interpret the correlogram to identify any significant autocorrelation patterns and their implications for forecasting

```{r}
# Compute the ACF
acf_result <- acf(data$Sales_transformed, lag.max = 20, plot = FALSE)

# Plot the ACF
plot(acf_result, main = "ACF of Transformed Sales Data")

```
In a correlogram, the dashed horizontal lines at approximately 0.2 and -0.2 serve as reference lines for zero autocorrelation. Autocorrelation Function (ACF) values exceeding these bounds suggest statistically significant correlations that may indicate meaningful patterns in the data.

For positive  correlation: 
 - If the ACF exceeds 0.2, it implies that past sales data at that specific lag significantly influences future sales data.For example, the ACF at lag 5 is 0.8,  indicates a strong positive correlation between sales data from 5 time periods ago and current sales.In other words, sales data from 5 periods back significantly influences the current sales data.This means that past sales data is correlated with future sales.Useful for forecasting,i.e, past sales can help predict future sales.
 - This information can guide forecasting models and decision-making.
 
For negative correlation:
 - If the ACF falls below -0.2, it suggests that past sales data at that lag significantly negatively affects future sales data.For instance, hypothetically if the ACF at lag 40 is -0.5, sales data from 40 time periods ago has a strong negative impact on current sales.Beyond this point, using past sales for forecasting becomes less effective.
 - Understanding such negative correlations can help identify potential areas for improvement or adjustment in business strategies.

Each vertical bar corresponds to a specific lag value and its height indicates the strength and direction of correlation at that lag.Most bars are above the horizontal line at ACF = 0, indicating positive autocorrelation at those lags.The first bar (at lag zero) is the tallest, suggesting strong positive autocorrelation since it’s correlated with itself.As the lag increases, ACF values generally decrease, indicating that sales data becomes less correlated with its (lagged) past values.

