---
title: "DESEASONALIZED"
author: "NEEMA NDANU"
date: "2024-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QUESTION 
### A manager of a national park collected data on the number of visitors (in thousands) who visited the park in each quarter of 2021 and 2022. The data includes:
Season Seasonal Index 2021 2022
Summer 162 51 54
Monsoon 62 28 32
Winter 87 41 45
Spring 89 36 43

### Tasks:1. Calculate the deseasonalized values for each quarter
```{r}
# Load the required libraries
library(tidyr)
library(dplyr)
library(ggplot2)

# Define the season
season_name <- c("Summer", "Monsoon", "Winter", "Spring")

# Define the season matrix
season_values <- matrix(c(51, 28, 41, 36, 54, 32, 45, 43), nrow = 4, ncol = 2, byrow = FALSE)

# Define the years
years <- c(2021, 2022)

# Define the seasonal index
seasonal_index <- c(162, 62, 87, 89)

# Assign column names to the season matrix
colnames(season_values) <- years

# Calculate the deseasonalized values
deseasonalized <- season_values / seasonal_index * 100

# Combine the years, season_name into a data frame
data <- data.frame(
  year = rep(years, each = length(season_name)),
  season_name = rep(season_name, times = length(years)),
  number_of_visit = as.vector(season_values),
  seasonal_index = rep(seasonal_index, times = length(years)),
  deseasonalized = as.vector(deseasonalized)
)

# Print the data frame
print(data)
```
### EXPLANATION
The deseasonalized values after the values without the seasonal effect.This helps in understanding the trend without the seasonal fluctuations.

### Compute the trend values for the deseasonalized data
```{r}
# Total number of observations
t <- nrow(data)

# Calculate the value of x for each observation
xt <- 2 * (1:t - 4.5)

# Sum of (xt * yt) and sum of x^2
yt <- data$deseasonalized
xtyt <- xt * yt
x_2 <- xt^2

# Calculate the B_0 and B_1 coefficients
B_1 <- sum(xtyt) / sum(x_2)
B_0 <- sum(yt) / t

# Print the results
print(B_1)
print(B_0)

# Find the trend values
Yt <- B_0 + B_1 * xt

# Create a data frame with the new values:
data_1 <- data.frame(
  year = rep(years, each = length(season_name)),
  season_name = rep(season_name, times = length(years)),
  number_of_visit = as.vector(season_values),
  seasonal_index = rep(seasonal_index, times = length(years)),
  deseasonalized = as.vector(deseasonalized),
  trend_values = Yt
)
print(data_1)

# Create the cyclic effect
data_2 <- data_1 %>%
  mutate(cyclic_effect = (deseasonalized / trend_values) * 100)
data_2
```
### EXPLANATION
The above trend values helps us to know the trend of the number of visits over time.

### Plot the quarterly number of visitors for 2021 and 2022, along with the deseasonalized values and trend values on the same graph
```{r}
# Create a time series object for adjusted index values
number_of_visit_ts <- ts(data_2$number_of_visit, start = c(2021, 1), frequency = 4)
deseasonalized_ts <- ts(data_2$deseasonalized, start = c(2021, 1), frequency = 4)
trend_values_ts <- ts(data_2$trend_values, start = c(2021, 1), frequency = 4)
cyclic_effect_ts <- ts(data_2$cyclic_effect, start = c(2021, 1), frequency = 4)

# Print the results
print(number_of_visit_ts)
print(deseasonalized_ts)
print(trend_values_ts)
print(cyclic_effect_ts)

# Create a new column combining year and season for plotting
data_2$season_year <- paste(data_2$year, data_2$season_name, sep = " ")

# Plot the data
library(ggplot2)
plot1 <- ggplot(data_2, aes(x = season_year, group = 1)) +
  geom_line(aes(y = number_of_visit, color = "Number of Visit")) +
  geom_line(aes(y = deseasonalized, color = "Deseasonalized")) +
  geom_line(aes(y = cyclic_effect, color = "Cyclic Effect")) +
  labs(title = "Seasonal Data Analysis", x = "Season (Year)", y = "Number of Visits") +
  scale_color_manual(values = c("Number of Visit" = "blue", "Deseasonalized" = "green", "Cyclic Effect" = "purple")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(limits = unique(data_2$season_year))

print(plot1)

```
### EXPLANATION
From the above graph, we can conclude the following:

1. The blue line indicates the number of visits with seasonal effects, showing peaks and troughs due to seasonal variations.
2. The green line indicates the number of visits after removing seasonal effects (deseasonalized data), such as changes due to summer, spring, monsoon, and winter.
3. The red line indicates the trend values of the deseasonalized data, showing an overall increasing trend in the number of visits over the years 2021 and 2022.
4. Comparing the three lines (number of visits, deseasonalized data, and trend), we can analyze how seasonality affects the overall number of visits. Even after removing the seasonality effects, the deseasonalized data still exhibits some peaks and troughs at specific times, such as peaks during the 2021 monsoon and winter, and similar peaks during the 2022 monsoon and winter.